\Large{Adversarial Training against Systematic Uncertainty}

### 2016-10-30 tex/paper.tex
No content change, just encoding
Some issue with encoding. No content changes.
http://stackoverflow.com/questions/21732903/git-sees-entire-file-as-one-
line-due-to-mac-line-endings
new abstract and intro, not complete

In high energy physics, systematic uncertainties represent our incomplete
knowledge in the theory of physical processes or in the properties of the
experimental detection apparatus. In effect, these uncertainties typically cause
variations in the conditional distributions of the data, thereby altering the
decisions and resulting efficacy of a classifier trained from it. To contain
this issue, we propose to repurpose adversarial training as a means to learn a
pivotal classifier which is invariant to variations of the nuisance parameters
describing the effects of systematic uncertainties. In particular, we show and
derive theoretical conditions under which classification optimality and
invariance with respect to systematics can be achieved, which we confirm
experimentally on a couple of illustrative examples. In machine learning, this
problem is closely related to those of domain adaptation and enforcing fairness
in classification. Following that line of work, the proposed method can be
regarded as generalization that also supports the continuous case, which can be
viewed as handling infinitely many domains or as enforcing fairness over
continuous features.

\end{abstract}

\maketitle

@@

The discovery of new particles or physical phenomena in high energy physics
experiments, like those at the Large Hadron Collider, requires the observation
of statistically significant deviations from the predictions of the Standard
Model, the current model of the known fundamental particles and their
interactions.  Typically, discovery requires satisfying the \textit{5-sigma
rule}, whereby the deviation must be at least five standard deviations from
predictions, i.e. having a p-value of $p \lesssim 3 \times 10^{-7}$. The
evaluation of the statistical significance must not only account for statistical
uncertainties due to random fluctuations, but also for
systematic uncertainties that represent our incomplete knowledge of the theory
of particle interactions or of the properties of the experimental detection
apparatus. Systematic uncertainties can alter the expected rate of signal and
background classes, e.g. their priors, as well as the distributions of features.
As such, the signal class true-positive rate and the background class
false-positive rate of a classifier can change with variations in the systematic
uncertainty thereby decreasing the sensitivity of an analysis. In this paper, we
present a new strategy for training a classifier, based on adversarial
training~\citep{goodfellow2014generative}, which aims at building a classifier
whose decision function is invariant under variations due to systematic
uncertainties.  In this way, uncertainties on predictions from simulations will
have a reduced impact when computing the significance of potential deviations
after application of the discriminant model and thereby increase the sensitivity
of the analysis to signals from potential new physics sources.

Given data $X$ and associated labels $Y$
taking values $y \in {\cal Y} = \{s,
b_{i=1...k} \}$ where $s$ is the signal class and $b_{i=1...k}$ are
the background classes, the probability density of the data
can be written as a mixture
\begin{equation}
p(X) = \pi_s p(X|Y=s) + \sum_{i=1}^{k} \pi_{b_i} p(X | Y=b_i),
\end{equation}
where $\pi_{y}$ represents the prior for the signal or a given
background, and $p(X|Y)$ is the conditional distribution of the data
for the signal or a given background. Systematic uncertainties can
affect this distribution in the following ways \citep{sinervo2003definition,bohm2010introduction}:
\begin{itemize}
\item Uncertainties on the priors $\pi_{y}$.  These arise from
  uncertainties on the rates we expect a given process to occur as
  predicted from theoretical calculations or from alternative
  measurements of the rate of a given class.

\item Uncertainties in the simulators of the physical interactions,
  giving rise to uncertainties on the distributions $p(X|Y)$.  These
  uncertainties are often determined by comparing the predictions of
  different simulations using differing models of the same physical
  process.

\item Uncertainties on the detection apparatus and its effect on the
  measurement of particle properties, giving rise to uncertainties on
  the distributions $p(X|Y)$.  For instance, this includes
  uncertainties on our knowledge of the mean and variance of the
  measured energy distribution of a particle.
\end{itemize}
In this paper, we will not address uncertainties of the first type
mentioned above (uncertainties on the priors).  Rather, we will focus
on mitigating the effect of systematic uncertainties that alter the
conditional distributions $p(X|Y)$. \glnote{Remove this comment?}

Systematic uncertainties are parameterized by nuisance parameters in the
statistical analysis of data (see e.g.,
\citep{bohm2010introduction,cowan2011asymptotic}).  Nuisance parameters often,
though not always, have a suitable prior coming from alternative measurements in
data.  For instance, the average measured energy of an electron in simulation
may be compared with those in data using a relatively pure sample of $Z$ boson
decays to electrons.  The average measured electron energy can not be measured
without uncertainty, and the measured uncertainty can be used in other analyses
to estimate the uncertainty deriving from our imperfect knowledge of the
electron's energy.  A nuisance parameter is assigned to the electron energy
uncertainty, which controls the variation of the electron energy in the
statistical analysis of data.  Nuisance parameters are typically, though not
always, constrained by a suitable prior, e.g. a Gaussian distribution with a
standard deviation equal to the measured uncertainty. \glnote{This paragraph
could be improved, it is not very clear what we want to say...}

@@

\section{Problem statement}
\label{sec:problem}

Let assume a probability space $(\Omega, {\cal F}, P)$, where $\Omega$ is a
sample space, ${\cal F}$ is a set of events and $P$ is a probability measure.
Let consider the multivariate random variables $X_z: \Omega \mapsto
\mathbb{R}^p$ and $Y: \Omega \mapsto {\cal Y}$, where $X_z$ denotes a dependence
of the functional $X$ on a nuisance parameter $Z$, whose values $z \in {\cal Z}$  define a
parameterized family of its systematic uncertainties. That is, $X_z$ and $Y$
induce together a joint probability distribution $p(X,Y|Z=z)$, where the
conditional denotes $X_z$. For training, let further assume a finite set $\{
x_i, y_i, z_i \}_{i=1}^N$ of realizations $X_{z_i}(\omega_i), Y(\omega_i)$, for
$\omega_i \in \Omega$ and known values $z_i$ of the nuisance parameter. Our goal
is to learn a score function $f(\cdot;\theta_f) : \mathbb{R}^p \mapsto {\cal S}$ of
parameters $\theta_f$ (e.g., a neural network-based probabilistic classifier) and minimizing  a loss ${\cal L}_f(\theta_f)$ (e.g.,
the cross-entropy). In addition, we require that $f(X_z ; \theta_f)$ should be
robust to the value $z$ of the nuisance parameter  -- which remains unknown at
test time. More specifically, we aim at building $f$ such that in the ideal case
\begin{equation}\label{eqn:criterion-true}
f(X_{z}(\omega) ; \theta_f) = f(X_{z^\prime}(\omega) ; \theta_f)
\end{equation} for all
samples $\omega \in \Omega$ and all $z, z^\prime$ pairs of values of the
nuisance parameter.

Since in general we do not have training tuples $(X_{z}(\omega),
X_{z^\prime}(\omega))$ (for the same unknown $\omega$), we propose instead to
solve the closely related problem of finding a predictive function $f$ such that
\begin{align}\label{eqn:criterion-measure}
    &P(\{ \omega | f(X_{z}(\omega) ; \theta_f) = s \} ) \nonumber \\
    &\indent = P( \{ \omega' | f(X_{z^\prime}(\omega') ; \theta_f) = s\})
\end{align}
for all values $s \in {\cal S}$ taken by the score function $f$. In words, we are looking for a predictive function $f$
which is a pivotal quantity \citep{degroot1986probability} with respect to the
nuisance parameter. That is, such that  the distribution of $f(X_z; \theta_f)$
is invariant with respect to the value $z$ of the nuisance. Note that a function
$f$ for which Eqn.~\ref{eqn:criterion-true} is true necessarily satisfies
Eqn.~\ref{eqn:criterion-measure}. In general, the converse is however not true,
since the sets of samples $\{ \omega | f(X_{z}(\omega); \theta_f) = s \}$ and $\{
\omega' | f(X_{z^\prime}(\omega'); \theta_f) = s \}$ do not need to be the same
for the equality to hold. In order to simplify notations, and as only
Eqn.~\ref{eqn:criterion-measure} is of direct interest in this work, we denote
from here on the pivotal quantity criterion as
\begin{equation}\label{eqn:criterion}
    p(f(X ; \theta_f) = s | z ) = p(f(X ; \theta_f) = s | z^\prime )
\end{equation}
for all $z,z^\prime \in  {\cal Z}$ and all values $s \in {\cal S}$ of $f(X ; \theta_f)$.

>>>>

Many inference problems involve data generation processes that are not uniquely specified or are uncertain in some way. 
In a scientific context, the presence of several plausible data generation processes is often
associated to the presence of systematic uncertainties. 
Robust inference is possible if it is based on a pivot -- a quantity whose
distribution is invariant to the unknown value of the nuisance parameters that parametrize a family of data generation processes. 
We introduce an adversarial training procedure to enforce the pivotal property on a predictive model, which 
allows one to tune the tradeoff between power and robustness. 
We demonstrate the effectiveness of this approach with a toy and real world example.

%Many inference problems involve data generation processes that are not uniquely specified or uncertain in some way. 
%In a scientific context, the presence of several plausible data generation processes is often
%associated to the presence of systematic uncertainty. 
%We consider the situation of a 
%family of data generation processes parametrized by nuisance parameters, which can
%be categorical or continuous. Robust inference is possible if it is based on a pivot -- a quantity whose
%distribution is invariant to the unknown value of these nuisance parameters. 
%We introduce an adversarial training procedure to enforce the pivotal property and
%allow one to tune the tradeoff between power and robustness.

%Our goal is to have an inferential procedure that is robust to the uncertainties in the data generation process. 
\end{abstract}

\maketitle
% ==============================================================================

\kcnote{ need some text like abstract here before diving into formal statement.}

@@

\section{Problem statement}
\label{sec:problem}
%
%Let assume a probability space $(\Omega, {\cal F}, P)$, where $\Omega$ is a
%sample space, ${\cal F}$ is a set of events and $P$ is a probability measure.
%Let consider the multivariate random variables $X_z: \Omega \mapsto
%\mathbb{R}^p$ and $Y: \Omega \mapsto {\cal Y}$, where $X_z$ denotes a dependence
%of the functional $X$ on a nuisance parameter $Z$, whose values $z \in {\cal Z}$  define a
%parameterized family of its systematic uncertainties. That is, $X_z$ and $Y$
%induce together a joint probability distribution $p(X,Y|Z=z)$, where the
%conditional denotes $X_z$. For training, let further assume a finite set $\{
%x_i, y_i, z_i \}_{i=1}^N$ of realizations $X_{z_i}(\omega_i), Y(\omega_i)$, for
%$\omega_i \in \Omega$ and known values $z_i$ of the nuisance parameter. Our goal
%is to learn a score function $f(\cdot;\theta_f) : \mathbb{R}^p \mapsto {\cal S}$ of
%parameters $\theta_f$ (e.g., a neural network-based probabilistic classifier) and minimizing  a loss ${\cal L}_f(\theta_f)$ (e.g.,
%the cross-entropy). In addition, we require that $f(X_z ; \theta_f)$ should be
%robust to the value $z$ of the nuisance parameter  -- which remains unknown at
%test time. More specifically, we aim at building $f$ such that in the ideal case
%\begin{equation}\label{eqn:criterion-true}
%f(X_{z}(\omega) ; \theta_f) = f(X_{z^\prime}(\omega) ; \theta_f)
%\end{equation} for all
%samples $\omega \in \Omega$ and all $z, z^\prime$ pairs of values of the
%nuisance parameter.
%
%Since in general we do not have training tuples $(X_{z}(\omega),
%X_{z^\prime}(\omega))$ (for the same unknown $\omega$),

Our goal
is to learn a score function $f(\cdot;\theta_f) : \mathbb{R}^p \mapsto {\cal S}$ of
parameters $\theta_f$ (e.g., a neural network-based probabilistic classifier) and minimizing  a loss ${\cal L}_f(\theta_f)$ (e.g.,
the cross-entropy). In addition, we require that the distribution of $f(X_z ; \theta_f)$ should be
robust to the value $z$ of the nuisance parameter  -- which remains unknown at
test time.

We propose  to
solve the  problem of finding a predictive function $f$ such that
\begin{equation}\label{eqn:criterion}
    p(f(X ; \theta_f) = s | z ) = p(f(X ; \theta_f) = s | z^\prime )
\end{equation}
for all $z,z^\prime \in  {\cal Z}$ and all values $s \in {\cal S}$ of $f(X ; \theta_f)$.
In words, we are looking for a predictive function $f$
which is a pivotal quantity \citep{degroot1986probability} with respect to the
nuisance parameter. That is, such that  the distribution of $f(X_z; \theta_f)$
is invariant with respect to the value $z$ of the nuisance.

%Note that a function
%$f$ for which Eqn.~\ref{eqn:criterion-true} is true necessarily satisfies
%Eqn.~\ref{eqn:criterion-measure}. In general, the converse is however not true,
%since the sets of samples $\{ \omega | f(X_{z}(\omega); \theta_f) = s \}$ and $\{
%\omega' | f(X_{z^\prime}(\omega'); \theta_f) = s \}$ do not need to be the same
%for the equality to hold. In order to simplify notations, and as only
%Eqn.~\ref{eqn:criterion-measure} is of direct interest in this work, we denote
%from here on the pivotal quantity criterion as

@@
